Burning sage: Reversing the curse of dimensionality in the visualization of high-dimensional data

### 2021-01-25 paper/paper.Rmd
Last tweaks for resubmission.

\textcolor{blue}{Here we are providing a solution to fix a long-standing problem in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections. Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions. This is a structural problem related purely to the geometry of projections from high-dimensional spaces. We show that it can be alleviated with a radial transformation on the low-dimensional space.} Figure \ref{fig:density} illustrates this crowding problem, as would be experienced when using a tour. Two-dimensional linear projections of data, generated by sampling uniformly within $p$-dimensional hyperspheres ($p=3, 10, 100$), are displayed as hexbin plots. Color indicates log count of the bin, with yellow mapping the highest counts. As $p$ increases the density concentrates in the center of the projection. \textcolor{blue}{(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons. Any linear projection is equivalent because of the rotation invariance of the distribution. The uniform sampling places focus on the space, points are spread evenly across the volume, and not a data distribution. Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the middle plot.)}

@@

\textcolor{blue}{Common solutions to overplotting include transparency and binned or density displays} [@chang]. \textcolor{blue}{These are designed to tackle large sample size, rather than dimension reduction problems. A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection. These approaches are still useful for large sample sizes, and could be applied after the transformation.}

\textcolor{red}{The underlying idea of the (radial) transformation, to reverse the curse, is to correct for the distortion in geometry that occurs when projecting from a high-dimensional space onto low dimensions. This is done by mapping equal volume in the full space onto equal area in a two-dimensional projection via a geometrical transformation of the projected data. When used with data that is uniformly distributed in the high-dimensional space, the transformation will result in a unform distribution in the low-dimensional display. In general, for \textit{any} underlying data distribution, we can instead think about the effect in terms of resolution: adjusting for the geometric piling effect means that the resolution depends on the projected volume. The observed projected distribution is representative of the original high-dimensional distribution in this sense.}

@@

\textcolor{red}{We may also wish to apply non-linear transformations before visualising the data or combining variables in a projection. One example would be to use a log transformation in the case of variable that has a skewed distribution. These types of transformations will generally change the distribution across the high-dimensional volume (and in particular also the position of the mean), and can thus make the sage display more useful for a wide range of observed data distributions.}

\textcolor{red}{The third common pre-processing would involve a dimension reduction step. Even with a relatively low number of dimensions, e.g. example with $p=10$, we see noticeable piling. However, we are often confronted with much higher dimensional datasets in which many of the variables are primarily adding noise. In this case it is advisable to first reduce the number of dimensions, for example using principal components analysis. Other dimension reduction techniques could be used. The sage tour would then be used to view the first few components, which would likely provide more insight about the data structures than is possible from only the first two. }

@@

```{r mouse-cluster, fig.cap = "Selected frames from using a tour of the mouse data with the default tour display (top), and the sage display with $\\gamma=3$ (bottom). Three selected clusters are highlighted in color, all other points are shown in grey. Using the sage display mitigates overplotting and provides a better understanding of cluster separation.", out.width="0.9\\textwidth", fig.align="center"}
library(magick)
dir <- here::here("../pngs/")
#dir <- here::here("pngs//")
frames <- c("001", "016", "038", "074")
draw_frames <- function(pngs) {
  res <- lapply(pngs, image_read)
  res <- do.call("c", res)
  image_append(res)
}

@@

Data were obtained from CT14HERA2 parton distribution function fits and describe the sensitivity of fit parameters to experimental measurements [@Wang:2018heo]. There are 28 parameters, and varying one at a time to move $\pm 1 \sigma$ away from the 'best fit point' (maximum likelihood estimate) provides our input variables, labelled X1-X56. Each of the 2808 observations corresponds to a physical observable and measures how the fit prediction changes along the 56 directions in parameter space. Points are grouped based on the underlying process in the experiment, which is mapped to color in the following. With the analysis of the distribution along these variables X1-X56 we can understand to what extent each experimental measurement provides new information for the global fit. For example, orthogonality between groups marks complementary constraints, and outlying points are considered as important for future fits, see discussion in @Cook:2018mvr.

>>>>

\textcolor{red}{Here we are providing a solution to fix a long-standing problem in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections. Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions. This is a structural problem related purely to the geometry of projections from high-dimensional spaces. We show that it can be alleviated with a radial transformation on the low-dimensional space.} Figure \ref{fig:density} illustrates this crowding problem, as would be experienced when using a tour. Two-dimensional linear projections of data, generated by sampling uniformly within $p$-dimensional hyperspheres ($p=3, 10, 100$), are displayed as hexbin plots. Color indicates log count of the bin, with yellow mapping the highest counts. As $p$ increases the density concentrates in the center of the projection. \textcolor{red}{(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons. Any linear projection is equivalent because of the rotation invariance of the distribution. The uniform sampling places focus on the space, points are spread evenly across the volume, and not a data distribution. Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the middle plot.)}

@@

\textcolor{red}{Common solutions to overplotting include transparency and binned or density displays} [@chang]. \textcolor{red}{These are designed to tackle large sample size, rather than dimension reduction problems. A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection. These approaches are still useful for large sample sizes, and could be applied after the transformation.}

\textcolor{red}{The underlying idea of the (radial) transformation, to reverse the curse, is to correct for the distortion in geometry that occurs when projecting from a high-dimensional space onto low dimensions. This is done by mapping equal volume in the full space onto equal area in a two-dimensional projection via a geometrical transformation of the projected data. When used with data that is uniformly distributed in the high-dimensional space, the transformation will result in a uniform distribution in the low-dimensional display. In general, for \textit{any} underlying data distribution, we can instead think about the effect in terms of resolution: adjusting for the geometric piling effect means that the resolution depends on the projected volume. The observed projected distribution is representative of the original high-dimensional distribution in this sense.}

@@

\textcolor{red}{We may also wish to apply non-linear transformations before visualising the data or combining variables in a projection. One example would be to use a log transformation in the case of a variable that has a skewed distribution. These types of transformations will generally change the distribution across the high-dimensional volume (and in particular also the position of the mean), and can thus make the sage display more useful for a wide range of observed data distributions.}

\textcolor{red}{The third common pre-processing would involve a dimension reduction step. Even with a relatively low number of dimensions, e.g. with $p=10$, we see noticeable piling. However, we are often confronted with much higher dimensional datasets in which many of the variables are primarily adding noise. In this case it is advisable to first reduce the number of dimensions, for example using principal components analysis. Other dimension reduction techniques could be used. The sage tour would then be used to view the first few components, which would likely provide more insight about the data structures than is possible from only the first two. }

@@

```{r mouse-cluster, fig.cap = "Selected frames from using a tour of the mouse data with the default tour display (top), and the sage display with $\\gamma=3$ (bottom). Three selected clusters are highlighted in color, all other points are shown in grey. Using the sage display mitigates overplotting and provides a better understanding of cluster separation.", out.width="0.9\\textwidth", fig.align="center"}
library(magick)
#dir <- here::here("../pngs/")
dir <- here::here("pngs//")
frames <- c("001", "016", "038", "074")
draw_frames <- function(pngs) {
  res <- lapply(pngs, image_read)
  res <- do.call("c", res)
  image_append(res)
}

@@

Data were obtained from CT14HERA2 parton distribution function fits and describe the sensitivity of fit parameters to experimental measurements [@Wang:2018heo]. There are 28 parameters, and varying one at a time to move $\pm 1 \sigma$ away from the 'best fit point' (maximum likelihood estimate) provides our input variables, labeled X1-X56. Each of the 2808 observations corresponds to a physical observable and measures how the fit prediction changes along the 56 directions in parameter space. Points are grouped based on the underlying process in the experiment, which is mapped to color in the following. With the analysis of the distribution along these variables X1-X56 we can understand to what extent each experimental measurement provides new information for the global fit. For example, orthogonality between groups marks complementary constraints, and outlying points are considered as important for future fits, see discussion in @Cook:2018mvr.

### 2021-01-25 paper/paper.tex
Last tweaks for resubmission.

\textcolor{blue}{Here we are providing a solution to fix a long-standing problem in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections. Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions. This is a structural problem related purely to the geometry of projections from high-dimensional spaces. We show that it can be alleviated with a radial transformation on the low-dimensional space.}
Figure \ref{fig:density} illustrates this crowding problem, as would be
experienced when using a tour. Two-dimensional linear projections of
data, generated by sampling uniformly within \(p\)-dimensional
hyperspheres (\(p=3, 10, 100\)), are displayed as hexbin plots. Color
indicates log count of the bin, with yellow mapping the highest counts.
As \(p\) increases the density concentrates in the center of the
projection.
\textcolor{blue}{(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons. Any linear projection is equivalent because of the rotation invariance of the distribution. The uniform sampling places focus on the space, points are spread evenly across the volume, and not a data distribution. Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the middle plot.)}

@@

\textcolor{blue}{Common solutions to overplotting include transparency and binned or density displays}
\citep{chang}.
\textcolor{blue}{These are designed to tackle large sample size, rather than dimension reduction problems. A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection. These approaches are still useful for large sample sizes, and could be applied after the transformation.}

@@

\textcolor{red}{We may also wish to apply non-linear transformations before visualising the data or combining variables in a projection. One example would be to use a log transformation in the case of variable that has a skewed distribution. These types of transformations will generally change the distribution across the high-dimensional volume (and in particular also the position of the mean), and can thus make the sage display more useful for a wide range of observed data distributions.}

>>>>

\textcolor{red}{Here we are providing a solution to fix a long-standing problem in the tour, where high-dimensional data is viewed by a movie of low-dimensional projections. Points crowding the center of the drawing canvas detract from reading the plots, even with relatively few dimensions. This is a structural problem related purely to the geometry of projections from high-dimensional spaces. We show that it can be alleviated with a radial transformation on the low-dimensional space.}
Figure \ref{fig:density} illustrates this crowding problem, as would be
experienced when using a tour. Two-dimensional linear projections of
data, generated by sampling uniformly within \(p\)-dimensional
hyperspheres (\(p=3, 10, 100\)), are displayed as hexbin plots. Color
indicates log count of the bin, with yellow mapping the highest counts.
As \(p\) increases the density concentrates in the center of the
projection.
\textcolor{red}{(Note that, sampling from a uniform in a sphere is used for this illustration for several reasons. Any linear projection is equivalent because of the rotation invariance of the distribution. The uniform sampling places focus on the space, points are spread evenly across the volume, and not a data distribution. Also, $p=10$ would already be considered high-dimensional, as seen by the piling observed in the middle plot.)}

@@

\textcolor{red}{Common solutions to overplotting include transparency and binned or density displays}
\citep{chang}.
\textcolor{red}{These are designed to tackle large sample size, rather than dimension reduction problems. A low-dimensional projection represented as a scatterplot, binned plot, or contour plot, will still suffer from data piling and reduced resolution near the center of the projection. These approaches are still useful for large sample sizes, and could be applied after the transformation.}

@@

\textcolor{red}{We may also wish to apply non-linear transformations before visualising the data or combining variables in a projection. One example would be to use a log transformation in the case of a variable that has a skewed distribution. These types of transformations will generally change the distribution across the high-dimensional volume (and in particular also the position of the mean), and can thus make the sage display more useful for a wide range of observed data distributions.}
